Supp mat: time series paper

### 2017-08-16 writing/manuscript_draft.Rmd
updated fig captions and figs

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. We only briefly discuss this approach as it has been described elsewhere. Essentially, we use a population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. population variability). As an example, we can take the following population model for population size $N$ at time $t$:
$$
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
$$
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase).

By simulating this model repetitively for different lengths of time, we can determine the minimum time series required. Here, statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend.

@@

```{r,echo=F,eval=T,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$) and (d) different levels of population variability ($\\sigma$). In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05.\\label{fig:theoretical_approach}'}
source("../scripts/theoretical-models/calculate_power.R")
source("../scripts/theoretical-models/simulate_pop.R")

@@

  #load command
  # Plotting commands
  par(mfrow=c(2,2),oma=c(1,2,0.5,0.5),mar=c(4,5,2.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='red')
      abline(h=0.8,lty=2,lwd=2,col='red')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3)
      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

```{r,echo=F,fig.cap='Distribution of the minimum time required in order to detect a significant trend (at the 0.05 level) in abundance given power of 0.8.\\label{fig:min_time_dist}'}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

counts <- table(pop_info$class,pop_info$min_time_for_power)
barplot(counts, main=" ",
        xlab=" ", ylab="", col=1:4,
        legend = rownames(counts),ylim=c(0,50),xlim=c(0,38),
        las=1,cex.names =1.2,cex.axis = 1.2,density=35,
        angle=c(20,40,60,80))

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}

>>>>

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. We only briefly discuss this approach as it has been described elsewhere. Essentially, we use a population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. population variability). As an example, we can take the following population model for population size $N$ at time $t$:
\begin{equation}
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase). It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs. \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).

We can determine the minimum time series required by simulating this model repetitively for different lengths of time. Here, statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

```{r,echo=F,eval=T,fig.cap='(a) Example of a simulated time series for 40 time periods. (b) Statistical power versus the simulated time series length. The horizontal, dashed line is the desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve the desired statistical power. (c) Minimum time required ($T_{min}$) for simulations with different values of the trend strength ($r$). (d) Minimum time required for different levels of population variability ($\\sigma$). In each case, the minimum time required is the minimum number of years to achieve 0.8 statistical power given a significance level of 0.05.\\label{fig:theoretical_approach}'}
source("../scripts/theoretical-models/calculate_power.R")
source("../scripts/theoretical-models/simulate_pop.R")

@@

  #load command
  # Plotting commands
  par(mfrow=c(2,2),oma=c(1.5,2,0.5,0.5),mar=c(3,5,2.5,0.5))
    matplot(simulations,type='l',col=rgb(0.5,0.5,0.5,0.5),lty=1,ylab='',xlab='',las=1,cex.axis=1.2)
      mtext(text = 'time' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'population size' ,side = 2,cex = 1.2,line=3.5)
      mtext(text = '(a)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

    plot(years_to_sample,power,type='l',lwd=2,las=1,ylab='',xlab='',cex.axis=1.2)
      abline(v=min_time_needed,lty=2,lwd=2,col='black')
      abline(h=0.8,lty=2,lwd=2,col='black')
      mtext(text = 'time sampled' ,side = 1,cex = 1.2,line=3)
      mtext(text = 'statistical power' ,side = 2,cex = 1.2,line=3)
      mtext(text = '(b)',side = 3,line = -1.2,adj=0.05,font = 2,cex=1.2)

@@

```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes. The minimum time required calculation corresponds to a significant level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

counts <- table(pop_info$class,pop_info$min_time_for_power)
barplot(counts, main=" ",
        xlab=" ", ylab="", col=gray.colors(4),
        ylim=c(0,50),xlim=c(0,40),
        las=1,cex.names =1.2,cex.axis = 1.2,#density=35,
        angle=c(20,40,60,80))
legend('topright',legend=rownames(counts),col = gray.colors(4),pch = 15,bty ="n")

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}

### 2017-08-16 writing/manuscript_draft.tex
updated fig captions and figs

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power for a particular question. This
approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\): \[
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\] where \(\epsilon\) is a normally-distributed random noise term with
mean \(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\)
is also the trend strength of the increase or decrease (i.e.~the rate of
increase).

By simulating this model repetitively for different lengths of time, we
can determine the minimum time series required. Here, statistical power
is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) example of simulated time series for 40 time periods and
(b) statistical power versus the simulated time series length. The
horizontal, dashed line is our desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve our
desired statistical power. (c) Minimum time required (\(T_{min}\)) for
simulations with different values of the trend strength (\(r\)) and (d)
different levels of population variability (\(\sigma\)). In each case,
the minimum time required is the minimum number of years to achieve 0.8
statistical power given a significance level of
0.05.\label{fig:theoretical_approach}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distribution of the minimum time required in order to detect a
significant trend (at the 0.05 level) in abundance given power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal
autocorrelation.\label{fig:correlates}}
\end{figure}

>>>>

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power for a particular question. This
approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\):

\begin{equation}
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth \(r\) is
also the trend strength of the increase or decrease (i.e.~the rate of
increase). It is important to note that any population model could be
subsituted for equation \ref{sim_model}, as we do in the supplementary
material (Figs. \ref{fig:min_time_dist_log_pop},
\ref{fig:shark_example}).

We can determine the minimum time series required by simulating this
model repetitively for different lengths of time. Here, statistical
power is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend. We also tested the effect of
varying both the significance level and statistical power (Fig.
\ref{fig:min_time_vs_alpha_beta}).

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{(a) Example of a simulated time series for 40 time periods. (b)
Statistical power versus the simulated time series length. The
horizontal, dashed line is the desired statistical power of 0.8. The
vertical, dashed line is the minimum time required to achieve the
desired statistical power. (c) Minimum time required (\(T_{min}\)) for
simulations with different values of the trend strength (\(r\)). (d)
Minimum time required for different levels of population variability
(\(\sigma\)). In each case, the minimum time required is the minimum
number of years to achieve 0.8 statistical power given a significance
level of 0.05.\label{fig:theoretical_approach}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{Distributions of the minimum time required for four different
biological classes. The minimum time required calculation corresponds to
a significant level of 0.05 and statistical power of
0.8.\label{fig:min_time_dist}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Minimum time required to estimate change in abundance
correlated with (a) trend strength (absolute value of slope coefficient
estimated from linear regression), (b) population variance (interannual
variability in population size), and (c) temporal lag-1
autocorrelation.\label{fig:correlates}}
\end{figure}
